---
title: "ExpectedPoints"
author: "Schuckers"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    theme: flatly
    highlight-style: github
execute:
  output: false
  warning: false
  message: false
---

## Overview
This document loads play-by-play data from the 2024 NFL season, removes garbage time (defined as plays where the win probability is outside the 5% to 95% range), and visualizes how well the Expected Points (EP) model calibrates to the actual next scoring events.

```{r}
#| label: setup
library(tidyverse)
library(nflfastR)
```

Load the 2024 NFL season data.

```{r}
#| label: load-data
pbp_2024 <- load_pbp(2024)
```


Filter the data and get the next score value

```{r}
#| label: filter-and-label
pbp_filtered <- pbp_2024 %>%
  # Remove extra points (EP assumes TDs are worth 7 flat) and plays missing data
  filter(
    season_type == "REG",
    !is.na(ep), 
    !is.na(posteam),
    !play_type %in% c("extra_point", "two_point_attempt")
  ) %>%
  group_by(game_id, game_half) %>%
  mutate(
    # Step A: Identify scoring events from a fixed (Home Team) perspective
    home_scoring_event = case_when(
      touchdown == 1 & td_team == home_team ~ 7,
      touchdown == 1 & td_team == away_team ~ -7,
      field_goal_result == "made" & posteam == home_team ~ 3,
      field_goal_result == "made" & posteam == away_team ~ -3,
      safety == 1 & defteam == home_team ~ 2,
      safety == 1 & defteam == away_team ~ -2,
      TRUE ~ NA_real_
    )
  ) %>%
  # Step B: Backfill the score to all preceding plays in the half
  tidyr::fill(home_scoring_event, .direction = "up") %>%
  mutate(
    # Step C: If no more scores happen in the half, the next score is 0
    home_scoring_event = coalesce(home_scoring_event, 0),
    
    # Step D: Convert back to the possession team's perspective
    pts_next_score = if_else(posteam == home_team, home_scoring_event, -home_scoring_event)
  ) %>%
  ungroup() %>%
  # Apply our garbage time filters
  filter(
    !is.na(wp),
    wp >= 0.05,
    wp <= 0.95
  )
```


### Calibration Plots

Next we are going to look at a calibration plot.  This plot is an important one in model prediction.  It plots
the predicted (or expected) values against the actual values.  (Not unlike a Q-Q plot.)  To simplify things,
it is often the case that we bin the predicted values.  Here they are binned into half point windows.  Here the binning is done by taking the predicted value bins and getting the mean of the actual points in the next score for those plays.  


```{r}
#| label: calibration-plot1
#| fig-width: 8
#| fig-height: 6

calibration_data <- pbp_filtered %>%
  mutate(ep_bin = round(ep * 2) / 2) %>% 
  group_by(ep_bin) %>%
  summarize(
    actual_pts = mean(pts_next_score),
    plays = n(),
    .groups = "drop"
  ) %>%
  # Filter out very small bins to reduce noise at extreme ends
  filter(plays >= 50) 

ggplot(calibration_data, aes(x = ep_bin, y = actual_pts)) +
  geom_point(aes(size = plays), alpha = 0.7, color = "#013369") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#D50A0A", linewidth = 1) +
  scale_x_continuous(breaks = seq(-4, 7, 1)) +
  scale_y_continuous(breaks = seq(-4, 7, 1)) +
  labs(
    title = "Expected Points Calibration (2024 NFL Season)",
    subtitle = "Actual Next Score vs. Expected Points (excluding garbage time)",
    x = "Expected Points (Binned)",
    y = "Average Actual Next Score",
    size = "Number of Plays",
    caption = "Data: nflfastR"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```
Why are calibration plots useful?


### Task 1

Redo the above using 0.25 points bins.  Hint change line 75 above.

```{r}

```

Redo the above using 2025 data

```{r}

```


So the Expected Points calculation from the Yurko et al paper using a multinomial logit to get the expected points.
What would happen if we used just the points for a given drive and we just used a regular regression model with the points per drive as the target/response.  Those are big changes to the model but let's try it and see what differences we find.


Sticking with the 2024 data.

```{r}
#| label: calculate-drive-scores
#| 
pbp_with_drive_score <- pbp_filtered %>%
  # Ensure the data is sorted chronologically
  arrange(game_id, play_id) %>%
  # 1. Create static post-play scores for home and away teams
  mutate(
    home_score_post = if_else(posteam == home_team, posteam_score_post, defteam_score_post),
    away_score_post = if_else(posteam == away_team, posteam_score_post, defteam_score_post)
  ) %>%
  # 2. Group by game and drive (filtering out pre-game/halftime missing drives)
  filter(!is.na(drive)) %>%
  group_by(game_id, drive) %>%
  mutate(
    # Calculate the points scored by each team strictly during this drive
    drive_home_pts = last(home_score_post, na_rm = TRUE) - first(total_home_score, na_rm = TRUE),
    drive_away_pts = last(away_score_post, na_rm = TRUE) - first(total_away_score, na_rm = TRUE),
    
    # Identify the team that began the drive with the ball
    drive_posteam = first(posteam, na_rm = TRUE),
    
    # 3. Calculate net score change relative to the initial possession team
    drive_score_change = if_else(
      drive_posteam == home_team,
      drive_home_pts - drive_away_pts,
      drive_away_pts - drive_home_pts
    )
  ) %>%
  ungroup() %>%
  # Select key columns to verify the results
  select(
    game_id, drive, play_id, desc, 
    drive_posteam, drive_home_pts, drive_away_pts, drive_score_change
  )

# View a sample of the output (e.g., a scoring drive)
pbp_with_drive_score %>%
  filter(drive_score_change != 0) %>%
  head(10) %>%
  knitr::kable()
```
```{r}


model_data <- pbp_filtered %>%
  arrange(game_id, play_id) %>%
  mutate(
    home_score_post = if_else(posteam == home_team, posteam_score_post, defteam_score_post),
    away_score_post = if_else(posteam == away_team, posteam_score_post, defteam_score_post)
  ) %>%
  filter(!is.na(drive)) %>%
  group_by(game_id, drive) %>%
  mutate(
    drive_home_pts = last(home_score_post, na_rm = TRUE) - first(total_home_score, na_rm = TRUE),
    drive_away_pts = last(away_score_post, na_rm = TRUE) - first(total_away_score, na_rm = TRUE),
    drive_posteam = first(posteam, na_rm = TRUE),
    drive_score_change = if_else(
      drive_posteam == home_team,
      drive_home_pts - drive_away_pts,
      drive_away_pts - drive_home_pts
    )
  ) %>%
  ungroup() %>%
  # Filter for plays with valid predictors
  filter(
    !is.na(down),
    !is.na(half_seconds_remaining),
    !is.na(yardline_100),
    !is.na(ydstogo),
    !is.na(goal_to_go),
    ydstogo > 0 # Prevent log(0) errors
  ) %>%
  mutate(
    # Create the custom predictors requested
    log_ydstogo = log(ydstogo),
    under_two_minutes = if_else(half_seconds_remaining <= 120, 1, 0),
    
    # Treat down as a categorical factor since the difference 
    # between 3rd and 4th down isn't the same as 1st and 2nd down
    down = as.factor(down) 
  )
```

```{r}
#| label: fit-model
ep_model <- lm(
  drive_score_change ~ down + half_seconds_remaining + yardline_100 + 
                       log_ydstogo + goal_to_go + under_two_minutes,
  data = model_data
)

# Generate predictions and append them to the dataset
model_data <- model_data %>%
  mutate(predicted_score_change = predict(ep_model, newdata = model_data))

# Quick look at the model summary
summary(ep_model)
```
Note that each of the predictors is significant, except maybe *goal_to_go*.  

How would you interpret each of these coefficients?

Let's look at the model calibration for our new model.

```{r}
#| label: calibration-plot2
#| fig-width: 8
#| fig-height: 6

calibration_data <- model_data %>%
  mutate(pred_bin = round(predicted_score_change * 2) / 2) %>%
  group_by(pred_bin) %>%
  summarize(
    actual_pts = mean(drive_score_change),
    plays = n(),
    .groups = "drop"
  ) %>%
  filter(plays >= 50) # Remove noisy extremes with small sample sizes

ggplot(calibration_data, aes(x = pred_bin, y = actual_pts)) +
  geom_point(aes(size = plays), alpha = 0.7, color = "#013369") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#D50A0A", linewidth = 1) +
  scale_x_continuous(breaks = seq(-4, 7, 1)) +
  scale_y_continuous(breaks = seq(-4, 7, 1)) +
  labs(
    title = "Custom Model Calibration (Linear Regression)",
    subtitle = "Actual Drive Score Change vs. Model Predicted Points",
    x = "Predicted Score Change (Binned)",
    y = "Average Actual Drive Score Change",
    size = "Number of Plays"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

From the above plot, we see that model is pretty well calibrated.

```{r}
#| label: compare-models-plot
#| fig-width: 8
#| fig-height: 6

comparison_data <- model_data %>%
  # Bin the standard nflfastR EP
  mutate(ep_bin = round(ep * 4) / 4) %>%
  group_by(ep_bin) %>%
  summarize(
    # Average our custom model's predictions within each standard EP bin
    avg_custom_pred = mean(predicted_score_change),
    plays = n(),
    .groups = "drop"
  ) %>%
  # Filter out extreme outliers with tiny sample sizes (like -4 EP)
  filter(plays >= 50)

ggplot(comparison_data, aes(x = ep_bin, y = avg_custom_pred)) +
  geom_point(aes(size = plays), alpha = 0.7, color = "#013369") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#D50A0A", linewidth = 1) +
  scale_x_continuous(breaks = seq(-4, 7, 1)) +
  scale_y_continuous(breaks = seq(-4, 7, 1)) +
  labs(
    title = "Model Comparison: nflfastR EP vs. Custom Linear EP",
    subtitle = "Comparing predicted values (2024 Season)",
    x = "Standard nflfastR EP (Binned)",
    y = "Average Custom Model Prediction",
    size = "Number of Plays"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```


### Task 2
Explain this plot.  Why are the blue dots not on the red dashed line?  Is that an indication of a bad model?



```{r}
#| label: plot-ep-curves
#| fig-width: 10
#| fig-height: 5

# Aggregate the data
plot_data <- model_data %>%
  # Filter to downs 1-4
  filter(down %in% c("1", "2", "3", "4")) %>%
  group_by(down, yardline_100) %>%
  summarize(
    nflfastR_EP = mean(ep, na.rm = TRUE),
    Custom_Linear_EP = mean(predicted_score_change, na.rm = TRUE),
    plays = n(),
    .groups = "drop"
  ) %>%
  # Filter out extremely rare yardlines to prevent jagged line spikes
  filter(plays >= 10) %>%
  # Pivot longer to make ggplot faceting easy
  pivot_longer(
    cols = c(nflfastR_EP, Custom_Linear_EP),
    names_to = "model_type",
    values_to = "ep_value"
  ) %>%
  mutate(
    model_type = if_else(model_type == "nflfastR_EP", "Standard nflfastR Model", "Custom Linear Model")
  )

# Create the visualization
ggplot(plot_data, aes(x = yardline_100, y = ep_value, color = down)) +
  # Using geom_smooth with a small span creates cleaner curves than raw geom_line
  geom_smooth(se = FALSE, span = 0.3, linewidth = 1.2) +
  facet_wrap(~ model_type, ncol = 2) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Expected Points by Down and Field Position",
    subtitle = "Comparing non-linear behavior (nflfastR) vs. strict linearity (Custom)",
    x = "Yards to Opponent Endzone (yardline_100)",
    y = "Expected Points (EP)",
    color = "Down"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r}
#| label: plot-ydstogo-curves
#| fig-width: 10
#| fig-height: 5

plot_data_ydstogo <- model_data %>%
  # Filter to downs 1-4 and realistic yards to go
  filter(
    down %in% c("1", "2", "3", "4"),
    ydstogo >= 1, 
    ydstogo <= 20
  ) %>%
  group_by(down, ydstogo) %>%
  summarize(
    nflfastR_EP = mean(ep, na.rm = TRUE),
    Custom_Linear_EP = mean(predicted_score_change, na.rm = TRUE),
    plays = n(),
    .groups = "drop"
  ) %>%
  # Ensure sufficient sample size per distance bucket
  filter(plays >= 10) %>%
  pivot_longer(
    cols = c(nflfastR_EP, Custom_Linear_EP),
    names_to = "model_type",
    values_to = "ep_value"
  ) %>%
  mutate(
    model_type = if_else(model_type == "nflfastR_EP", "Standard nflfastR Model", "Custom Linear Model")
  )

# Plotting EP vs Yards To Go
ggplot(plot_data_ydstogo, aes(x = ydstogo, y = ep_value, color = down)) +
  geom_smooth(se = FALSE, span = 0.4, linewidth = 1.2) +
  facet_wrap(~ model_type, ncol = 2) +
  scale_x_continuous(breaks = seq(1, 20, 2)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Expected Points by Down and Yards to Go",
    subtitle = "Comparing complex interactions (nflfastR) vs. log-transformed distance (Custom)",
    x = "Yards to Go (ydstogo)",
    y = "Expected Points (EP)",
    color = "Down"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

Well that is an interesting plot, isn't it.  What do we think is going on here?


### Task 3

Let's change what we filter.  Removing what we already do, let's also filter out any plays in the 4th quarter where the win probability is less than 0.1 OR more than 0.90.  How does that change our model and our plots?

```{r}

```


### Task 4

Now rerun all of this with data from 2022 to 2025.  What, if anything, changes?  Use original filtering from win probabilities.

```{r}


```




