---
title: "Explaining FBS Win Probability via Post-Game Box Scores"
author: "Schuckers"
date: "2026-08-31"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: show
    df-print: paged
execute:
  warning: false
  message: false
---

### Executive Summary

This analysis constructs an explanatory classification workflow analyzing FBS vs. FBS college football games from **2020 to 2025**. Using real game-level box score statistics sourced via `cfbfastR`, we model the post-game probability of winning ($P(\text{Win} = 1)$) based on game-level performance differentials.

### 1. Setup & Data Loading

We load the required R packages and pull real post-game team statistics using `cfbfastR`. 

```{r echo=FALSE}
Sys.setenv(CFBD_API_KEY = "MullWLpxkoKVq3XEFcHdC+RiYq1VtVUbuTdPz9v954k7/YmfI0m05GVO2jn+k7aj")
```

```{r}

library(cfbfastR)
library(tidyverse)
library(rsample)
library(broom)
library(yardstick)
library(knitr)
library(scales)
library(recipes)
```

*Decision: how to structure data?*

Each team separately, OR each team as a difference, OR home and away on same line 

Below I started by looking at the differences.  Note that
the argument *rows_per_team* seems to be misnamed and should be *rows_per_game*.

```{r}
seasons <- 2020:2025
weeks <- 1:15

# Container list for weekly box score stats
stats_list <- list()

# Loop through each season and week to collect game team stats
for (yr in seasons) {
  for (wk in weeks) {
    print(paste0("Attempting week " ,wk, " of ",yr))
    stats <- tryCatch({
      cfbfastR::cfbd_game_team_stats(
        year = yr,
        week = wk,
        division = "fbs",
        rows_per_team = 1
      ) %>%
        mutate(season = as.character(yr), week_num = wk)
    }, error = function(e) {
      NULL
    })
    
    if (!is.null(stats) && nrow(stats) > 0) {
      stats_list[[paste(yr, wk, sep = "_")]] <- stats
    }
  }
}

raw_team_stats <- bind_rows(stats_list)



glimpse(raw_team_stats)
```





### Data Filter and feature engineering

```{r}
#function to change time into seconds
parse_time_sec <- function(time_str) {
  parts <- strsplit(as.character(time_str), ":")[[1]]
  if (length(parts) == 2) {
    return(as.numeric(parts[1]) * 60 + as.numeric(parts[2]))
  }
  return(1800)
}

# clear the data
cfb_cleaned <- raw_team_stats %>%
  filter(!is.na(points), !is.na(points_allowed), points != points_allowed) %>%
  mutate(
    win = if_else(as.numeric(points) > as.numeric(points_allowed), 1, 0),
    win_factor = factor(win, levels = c(1, 0), labels = c("Win", "Loss")),
    season = factor(season)
  ) %>%
  separate(third_down_eff, c("td_comp", "td_att"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(third_down_eff_allowed, c("td_comp_opp", "td_att_opp"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(total_penalties_yards, c("pen_cnt", "pen_yds"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(total_penalties_yards_allowed, c("pen_cnt_opp", "pen_yds_opp"), sep = "-", convert = TRUE, fill = "right") %>%
  rowwise() %>%
  mutate(
    top_sec = parse_time_sec(possession_time),
    top_sec_opp = parse_time_sec(possession_time_allowed)
  ) %>%
  ungroup() %>%
  mutate(
    across(c(total_yards, net_passing_yards, rushing_yards, turnovers, 
             first_downs, turnovers_allowed, total_yards_allowed, 
             net_passing_yards_allowed, rushing_yards_allowed, first_downs_allowed,
             pen_yds, pen_yds_opp), as.numeric),
    
    third_down_pct     = if_else(td_att > 0, td_comp / td_att, 0),
    third_down_pct_opp = if_else(td_att_opp > 0, td_comp_opp / td_att_opp, 0),
    
    # Differential Features
    turnover_margin    = turnovers_allowed - turnovers,
    pass_yards_diff    = net_passing_yards - net_passing_yards_allowed,
    rush_yards_diff    = rushing_yards - rushing_yards_allowed,
    first_downs_diff   = first_downs - first_downs_allowed,
    third_down_pct_diff= third_down_pct - third_down_pct_opp,
    penalty_yards_diff = pen_yds - pen_yds_opp,
    top_seconds_diff   = top_sec - top_sec_opp
  ) %>%
  select(game_id, season, week_num, win, win_factor, home_away,turnover_margin, pass_yards_diff, 
         rush_yards_diff, first_downs_diff, third_down_pct_diff, 
         penalty_yards_diff, top_seconds_diff) %>%
  drop_na()

cfb_cleaned <- cfb_cleaned %>% 
  slice(which(row_number() %% 2 == 0))

head(cfb_cleaned) %>% kable(digits = 2)
```


```{r}
set.seed(2026)

game_split <- initial_split(cfb_cleaned, prop = 0.80, strata = season)
train_data <- training(game_split)
test_data  <- testing(game_split)

# Train vs. Test proportions by season
season_split_summary <- train_data %>%
  count(season, name = "Train_Games") %>%
  inner_join(test_data %>% count(season, name = "Test_Games"), by = "season") %>%
  mutate(
    Total_Games = Train_Games + Test_Games,
    Train_Pct = round(Train_Games / Total_Games * 100, 1)
  )

kable(season_split_summary, caption = "Season-Stratified Train/Test Split Distribution (2020–2025)")
```

```{r}


predictors <- c("turnover_margin", "pass_yards_diff", "rush_yards_diff", 
                "first_downs_diff", "third_down_pct_diff", 
                "penalty_yards_diff", "top_seconds_diff")

# Create recipe to normalize features on the training set
cfb_recipe <- recipe(win ~ ., data = train_data %>% select(win,home_away, all_of(predictors))) %>%
  step_normalize(all_of(predictors)) %>%
  prep()

# Bake standardized training and test datasets (guarantees identical numeric types)
train_scaled <- bake(cfb_recipe, new_data = NULL)
test_scaled  <- bake(cfb_recipe, new_data = test_data)

# Fit Logistic Regression Model
win_model_z <- glm(win ~ ., data = train_scaled, family = binomial(link = "logit"))

# Feature Importance Ranking
feature_importance <- tidy(win_model_z, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Predictor = term,
    Importance_z = abs(statistic),
    Odds_Ratio_1SD = exp(estimate),
    OR_Low = exp(conf.low),
    OR_High = exp(conf.high)
  ) %>%
  arrange(desc(Importance_z))

kable(
  feature_importance %>% 
    select(Predictor, `Std Beta` = estimate, `Odds Ratio (1 SD)` = Odds_Ratio_1SD, `z-statistic` = statistic, `Importance (|z|)` = Importance_z, `p-value` = p.value),
  digits = 3,
  caption = "Post-Game Feature Importance Ranked by Standardized |z|-score (with Home Advantage)"
)

# Plot Feature Importance
ggplot(feature_importance, aes(x = reorder(Predictor, Importance_z), y = Importance_z, fill = Importance_z)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  labs(
    title = "FBS Win Probability: Feature Importance Ranking (Incl. Home Team)",
    subtitle = "Standardized Wald z-statistic magnitude across season-stratified training data",
    x = "Model Feature",
    y = "Feature Importance (|z-score|)"
  ) +
  theme_minimal()
```

```{r}
# Predict probabilities on the prepared test set
test_preds <- test_data %>%
  mutate(
    pred_prob = predict(win_model_z, newdata = test_scaled, type = "response"),
    pred_class = factor(if_else(pred_prob >= 0.5, 1, 0), levels = c("1", "0")),
    actual_class = factor(win, levels = c("1", "0"))
  )

overall_metrics <- tibble(
  Metric = c("Accuracy", "ROC AUC", "Log Loss", "Brier Score"),
  `Test Score` = c(
    accuracy_vec(test_preds$actual_class, test_preds$pred_class),
    roc_auc_vec(test_preds$actual_class, test_preds$pred_prob),
    mn_log_loss_vec(test_preds$actual_class, test_preds$pred_prob),
    mean((test_preds$win - test_preds$pred_prob)^2)
  )
)

kable(overall_metrics, digits = 4, caption = "Out-of-Sample Performance on Stratified Test Data")
```

### Calibration plots

#### Calibration plot all seasons

```{r}


# Compute binned observed win rates across all test seasons combined
calibration_data_overall <- test_preds %>%
  mutate(
    prob_bin = cut(pred_prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(pred_prob),
    actual_win_rate = mean(win),
    game_count = n(),
    .groups = "drop"
  ) %>%
  drop_na()

# Plot aggregate calibration curve
ggplot(calibration_data_overall, aes(x = mean_pred, y = actual_win_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  geom_line(color = "royalblue", linewidth = 1) +
  geom_point(aes(size = game_count), color = "royalblue", alpha = 0.8) +
  scale_x_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(3, 8), labels = comma) +
  labs(
    title = "Overall Win Probability Calibration (Combined Test Set)",
    subtitle = "Dashed line = perfect calibration; Points = binned mean predictions vs. observed win rate",
    x = "Mean Predicted Win Probability",
    y = "Observed Win Rate",
    size = "Test Sample Size"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(face = "bold")
  )
```


#### Calibration plots by season

```{r}


# Compute binned observed win rates per season
calibration_data <- test_preds %>%
  mutate(
    prob_bin = cut(pred_prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
  ) %>%
  group_by(season, prob_bin) %>%
  summarise(
    mean_pred = mean(pred_prob),
    actual_win_rate = mean(win),
    game_count = n(),
    .groups = "drop"
  ) %>%
  drop_na()

# Plot calibration curves faceted by season
ggplot(calibration_data, aes(x = mean_pred, y = actual_win_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  geom_line(color = "royalblue", linewidth = 1) +
  geom_point(aes(size = game_count), color = "royalblue", alpha = 0.8) +
  facet_wrap(~ season, ncol = 3) +
  scale_x_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    title = "Win Probability Model Calibration by Season (2020–2025 Test Set)",
    subtitle = "Dashed line = perfect calibration; Points = binned mean predictions vs observed wins",
    x = "Mean Predicted Win Probability",
    y = "Observed Win Rate",
    size = "Test Games"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11)
  )
```

## Change the data structure to single team

```{r}
#seasons <- 2024:2025
#weeks <- 1:15

stats_list <- list()

for (yr in seasons) {
  for (wk in weeks) {
    print(paste0("Attempting week " ,wk, " of ",yr))
    stats <- tryCatch({
      cfbfastR::cfbd_game_team_stats(
        year = yr,
        week = wk,
        division = "fbs",
        rows_per_team = 2
      ) %>%
        mutate(season = as.character(yr), week_num = wk)
    }, error = function(e) {
      NULL
    })
    
    if (!is.null(stats) && nrow(stats) > 0) {
      stats_list[[paste(yr, wk, sep = "_")]] <- stats
    }
  }
}
```

### Data Filtering and Feature Engineering

```{r}
parse_time_sec <- function(time_str) {
  parts <- strsplit(as.character(time_str), ":")[[1]]
  if (length(parts) == 2) {
    return(as.numeric(parts[1]) * 60 + as.numeric(parts[2]))
  }
  return(1800)
}

cfb_cleaned <- raw_team_stats %>%
  filter(!is.na(points), !is.na(points_allowed), points != points_allowed) %>%
  mutate(
    win = if_else(as.numeric(points) > as.numeric(points_allowed), 1, 0),
    win_factor = factor(win, levels = c(1, 0), labels = c("Win", "Loss")),
    season = factor(season),
    is_home = if_else(str_to_lower(home_away) == "home", 1, 0)
  ) %>%
  separate(third_down_eff, c("td_comp", "td_att"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(third_down_eff_allowed, c("td_comp_opp", "td_att_opp"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(total_penalties_yards, c("pen_cnt", "pen_yds"), sep = "-", convert = TRUE, fill = "right") %>%
  separate(total_penalties_yards_allowed, c("pen_cnt_opp", "pen_yds_opp"), sep = "-", convert = TRUE, fill = "right") %>%
  rowwise() %>%
  mutate(
    top_sec = parse_time_sec(possession_time),
    top_sec_opp = parse_time_sec(possession_time_allowed)
  ) %>%
  ungroup() %>%
  mutate(
    across(c(total_yards, net_passing_yards, rushing_yards, turnovers, 
             first_downs, turnovers_allowed, total_yards_allowed, 
             net_passing_yards_allowed, rushing_yards_allowed, first_downs_allowed,
             pen_yds, pen_yds_opp), as.numeric),
    
    third_down_pct     = if_else(td_att > 0, td_comp / td_att, 0),
    third_down_pct_opp = if_else(td_att_opp > 0, td_comp_opp / td_att_opp, 0)
  ) %>%
  select(
    game_id, season, week_num, win, win_factor, is_home,
    # Individual Team Performance Features
    turnovers, turnovers_allowed,
    rushing_yards, rushing_yards_allowed,
    net_passing_yards, net_passing_yards_allowed,
    first_downs, first_downs_allowed,
    third_down_pct, third_down_pct_opp,
    pen_yds, top_sec
  ) %>%
  drop_na()

cfb_cleaned <- cfb_cleaned %>% 
  slice(which(row_number() %% 2 == 0))

head(cfb_cleaned) %>% kable(digits = 2)
```

### Split Data into Train/Test stratified by season

```{r}
set.seed(2026)

game_split <- initial_split(cfb_cleaned, prop = 0.80, strata = season)
train_data <- training(game_split)
test_data  <- testing(game_split)

season_split_summary <- train_data %>%
  count(season, name = "Train_Games") %>%
  inner_join(test_data %>% count(season, name = "Test_Games"), by = "season") %>%
  mutate(
    Total_Games = Train_Games + Test_Games,
    Train_Pct = round(Train_Games / Total_Games * 100, 1)
  )

kable(season_split_summary, caption = "Season-Stratified Train/Test Split Distribution (2020–2025)")
```

### Rescale the predictors

```{r}
predictors <- c(
  "is_home", "turnovers", "turnovers_allowed",
  "rushing_yards", "rushing_yards_allowed",
  "net_passing_yards", "net_passing_yards_allowed",
  "first_downs", "first_downs_allowed",
  "third_down_pct", "third_down_pct_opp",
  "pen_yds", "top_sec"
)

# Preprocessing recipe: Normalize numeric features on the training set
cfb_recipe <- recipe(win ~ ., data = train_data %>% select(win, all_of(predictors))) %>%
  step_normalize(all_of(predictors)) %>%
  prep()

# Prepare normalized train and test sets
train_scaled <- bake(cfb_recipe, new_data = NULL)
test_scaled  <- bake(cfb_recipe, new_data = test_data)

# Fit Logistic Regression Model on Individual Features
win_model_z <- glm(win ~ ., data = train_scaled, family = binomial(link = "logit"))

# Feature Importance Ranking
feature_importance <- tidy(win_model_z, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    Predictor = term,
    Importance_z = abs(statistic),
    Odds_Ratio_1SD = exp(estimate),
    OR_Low = exp(conf.low),
    OR_High = exp(conf.high)
  ) %>%
  arrange(desc(Importance_z))

kable(
  feature_importance %>% 
    select(Predictor, `Std Beta` = estimate, `Odds Ratio (1 SD)` = Odds_Ratio_1SD, `z-statistic` = statistic, `Importance (|z|)` = Importance_z, `p-value` = p.value),
  digits = 3,
  caption = "Individual Team Feature Importance Ranked by Standardized |z|-score"
)

# Feature Importance Visualization
ggplot(feature_importance, aes(x = reorder(Predictor, Importance_z), y = Importance_z, fill = Importance_z)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_c(option = "mako", direction = -1) +
  labs(
    title = "FBS Win Probability: Individual Team Feature Importance",
    subtitle = "Standardized Wald z-statistic magnitude (No net differential metrics)",
    x = "Individual Team Metric",
    y = "Feature Importance (|z-score|)"
  ) +
  theme_minimal()
```

Create predictions

```{r}
test_preds <- test_data %>%
  mutate(
    pred_prob = predict(win_model_z, newdata = test_scaled, type = "response"),
    pred_class = factor(if_else(pred_prob >= 0.5, 1, 0), levels = c("1", "0")),
    actual_class = factor(win, levels = c("1", "0"))
  )

overall_metrics <- tibble(
  Metric = c("Accuracy", "ROC AUC", "Log Loss", "Brier Score"),
  `Test Score` = c(
    accuracy_vec(test_preds$actual_class, test_preds$pred_class),
    roc_auc_vec(test_preds$actual_class, test_preds$pred_prob),
    mn_log_loss_vec(test_preds$actual_class, test_preds$pred_prob),
    mean((test_preds$win - test_preds$pred_prob)^2)
  )
)

kable(overall_metrics, digits = 4, caption = "Out-of-Sample Performance on Individual Team Feature Model")
```

#### Calibration in Test Set

```{r}
calibration_data_overall <- test_preds %>%
  mutate(
    prob_bin = cut(pred_prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
  ) %>%
  group_by(prob_bin) %>%
  summarise(
    mean_pred = mean(pred_prob),
    actual_win_rate = mean(win),
    game_count = n(),
    .groups = "drop"
  ) %>%
  drop_na()

ggplot(calibration_data_overall, aes(x = mean_pred, y = actual_win_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  geom_line(color = "royalblue", linewidth = 1) +
  geom_point(aes(size = game_count), color = "royalblue", alpha = 0.8) +
  scale_x_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(3, 8), labels = comma) +
  labs(
    title = "Overall Win Probability Calibration (Individual Team Features)",
    subtitle = "Combined test set across all seasons (2020–2025)",
    x = "Mean Predicted Win Probability",
    y = "Observed Win Rate",
    size = "Test Games"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(face = "bold")
  )
```


#### Calibration plots by season

```{r}


# Compute binned observed win rates per season
calibration_data <- test_preds %>%
  mutate(
    prob_bin = cut(pred_prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)
  ) %>%
  group_by(season, prob_bin) %>%
  summarise(
    mean_pred = mean(pred_prob),
    actual_win_rate = mean(win),
    game_count = n(),
    .groups = "drop"
  ) %>%
  drop_na()

# Plot calibration curves faceted by season
ggplot(calibration_data, aes(x = mean_pred, y = actual_win_rate)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  geom_line(color = "royalblue", linewidth = 1) +
  geom_point(aes(size = game_count), color = "royalblue", alpha = 0.8) +
  facet_wrap(~ season, ncol = 3) +
  scale_x_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    title = "Win Probability Model Calibration by Season (2020–2025 Test Set)",
    subtitle = "Dashed line = perfect calibration; Points = binned mean predictions vs observed wins",
    x = "Mean Predicted Win Probability",
    y = "Observed Win Rate",
    size = "Test Games"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 11)
  )
```