---
title: "Repeatability in the NFL"
author: "Schuckers"
format: 
  html:
    toc: true
    theme: default
execute:
  output: false
  warning: false
  message: false
---


## Environment Setup & Required Libraries

```{r}
# Uncomment if packages are not yet installed:
# install.packages(c("tidyverse", "nflfastR", "nflplotR", "gt", "knitr"))
library(tidyverse)
library(nflfastR)
library(nflplotR)
library(gt)
library(knitr)

# Set global option to avoid scientific notation in printouts
options(scipen = 999)
```

```{r}
# Fetch 2022 through 2025 regular season play-by-play data
seasons_vector <- 2022:2025

pbp_multi <- nflfastR::load_pbp(seasons_vector) |>
  filter(
    season_type == "REG",
    !is.na(posteam),
    play_type %in% c("pass", "run") | pass == 1 | rush == 1,
    !is.na(epa)
  )

# Display dataset size across all 4 seasons
cat("Total analyzed scrimmage plays (2022–2025):", nrow(pbp_multi), "\n")
```

```{r}
# Weekly Offensive Metrics
weekly_offense <- pbp_multi |>
  group_by(season, week, posteam) |>
  summarize(
    off_plays = n(),
    off_epa_per_play = mean(epa, na.rm = TRUE),
    off_pass_epa = mean(epa[pass == 1], na.rm = TRUE),
    off_rush_epa = mean(epa[rush == 1], na.rm = TRUE),
    off_success_rate = mean(success, na.rm = TRUE),
    pass_rate = mean(pass == 1, na.rm = TRUE),
    complete_pass_pct=mean(complete_pass,na.rm=TRUE),
    .groups = "drop"
  )

# Weekly Defensive Metrics
weekly_defense <- pbp_multi |>
  group_by(season, week, defteam) |>
  summarize(
    def_plays = n(),
    def_epa_allowed = mean(epa, na.rm = TRUE),
    def_success_rate_allowed = mean(success, na.rm = TRUE),
    .groups = "drop"
  )

# Master Panel Dataset (Weekly Team Summary)
weekly_team_panel <- weekly_offense |>
  inner_join(
    weekly_defense, 
    by = c("season" = "season", "week" = "week", "posteam" = "defteam")
  ) |>
  rename(team = posteam) |>
  arrange(team, season, week)

# Preview first 10 rows of weekly panel data
head(weekly_team_panel, 10) |> kable()
```

Next we are going to break the data into half seasons at week 9
```{r}
halfseason_team_panel <- weekly_team_panel %>%
  mutate(half_season  = if_else(week <= 9, "H1", "H2")) |>
  group_by(season, team, half_season) |>
  summarize(
    mean_off_plays = mean(off_plays,na.rm=TRUE),
    mean_off_epa_per_play = mean(off_epa_per_play, na.rm = TRUE),
    mean_off_pass_epa = mean(off_pass_epa, na.rm = TRUE),
    mean_off_rush_epa = mean(off_rush_epa, na.rm = TRUE),
    mean_off_success_rate = mean(off_success_rate, na.rm = TRUE),
    mean_pass_rate = mean(pass_rate, na.rm = TRUE),
    mean_complete_pass_pct=mean(complete_pass_pct,na.rm=TRUE),
    mean_def_plays = mean(def_plays,na.rm=TRUE),
    mean_def_epa_allowed = mean(def_epa_allowed, na.rm = TRUE),
    mean_def_success_rate_allowed = mean(def_success_rate_allowed, na.rm = TRUE)
  )
         
         
         
```
#### Task 
Explore these data and try to surmise which metrics are good ones.




### Correlations within the 2025 season
We'll start by doing 2025 season and doing some correlations.

```{r}
#  Load 2025 NFL play-by-play data
pbp_raw <- nflfastR::load_pbp(2025)

# Filter to regular season scrimmage plays
pbp_clean <- pbp_raw %>%
  filter(
    season_type == "REG",
    play == 1,
    !is.na(epa),
    !is.na(posteam),
    play_type %in% c("pass", "run")
  ) %>%
  mutate(
    # Create split-half indicator (Weeks 1-9 vs Weeks 10-18)
    half_season  = if_else(week <= 9, "H1", "H2"),
    is_explosive = if_else(yards_gained >= 15, 1, 0),
    is_success   = if_else(epa > 0, 1, 0)
  )
```

Turn the play by play data into some drive level data
```{r}
# Drive-Level Metrics (drive_epa, turnovers_per_drive)
drive_summary <- pbp_clean %>%
  group_by(game_id, drive, posteam, half_season) %>%
  summarize(
    drive_total_epa = sum(epa, na.rm = TRUE),
    drive_turnovers = sum(interception == 1 | fumble_lost == 1, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(posteam, half_season) %>%
  summarize(
    drive_epa           = mean(drive_total_epa, na.rm = TRUE),
    turnovers_per_drive = mean(drive_turnovers, na.rm = TRUE),
    .groups = "drop"
  )

# Play-Level Metrics
play_summary <- pbp_clean %>%
  group_by(posteam, half_season) %>%
  summarize(
    mean_epa       = mean(epa, na.rm = TRUE),
    success_rate   = mean(is_success, na.rm = TRUE),
    pass_epa       = mean(epa[play_type == "pass"], na.rm = TRUE),
    rush_epa       = mean(epa[play_type == "run"], na.rm = TRUE),
    yards_per_play = mean(yards_gained, na.rm = TRUE),
    completion_pct = mean(complete_pass[pass_attempt == 1], na.rm = TRUE),
    .groups = "drop"
  )

# Combine metrics
metrics_combined <- play_summary %>%
  left_join(drive_summary, by = c("posteam", "half_season"))

head(metrics_combined)
```



```{r}
# Combine all the metrics
metrics_combined <- play_summary %>%
  left_join(drive_summary, by = c("posteam", "half_season"))

head(metrics_combined)
```

Reshape the data to get the correlations.
```{r}
# Reshape data to wide format: metric_H1 vs metric_H2
metrics_wide <- metrics_combined %>%
  pivot_wider(
    names_from = half_season,
    values_from = c(
      mean_epa, success_rate, pass_epa, rush_epa,
      yards_per_play, drive_epa, completion_pct, turnovers_per_drive
    )
  )

metric_list <- c(
  "mean_epa", "success_rate", "pass_epa", "rush_epa",
  "yards_per_play", "drive_epa", "completion_pct", "turnovers_per_drive"
)
```

```{r}
# Compute Pearson r, p-value, and R-squared
correlation_results <- tibble(
  metric = metric_list,
  r = map_dbl(metric_list, ~ {
    col_h1 <- paste0(.x, "_H1")
    col_h2 <- paste0(.x, "_H2")
    cor(metrics_wide[[col_h1]], metrics_wide[[col_h2]], use = "complete.obs")
  }),
  p_value = map_dbl(metric_list, ~ {
    col_h1 <- paste0(.x, "_H1")
    col_h2 <- paste0(.x, "_H2")
    cor.test(metrics_wide[[col_h1]], metrics_wide[[col_h2]])$p.value
  }),
  r_squared = r^2
) %>%
  arrange(desc(r))
```


```{r}
# Render Summary Table
correlation_results %>%
  gt() %>%
  tab_header(
    title = "2025 NFL Offense Split-Half Reliability",
    subtitle = "Pearson Correlation (r) between Weeks 1-9 (H1) and Weeks 10-18 (H2)"
  ) %>%
  cols_label(
    metric = "Offensive Metric",
    r = "Pearson r",
    p_value = "p-value",
    r_squared = "R²"
  ) %>%
  fmt_number(columns = c(r, p_value, r_squared), decimals = 3)
```

```{r}
ggplot(correlation_results, aes(x = reorder(metric, r), y = r, fill = r)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = sprintf("r = %.2f", r)), hjust = -0.15, size = 3.8) +
  coord_flip() +
  scale_y_continuous(limits = c(-0.2, 1.0), breaks = seq(-0.2, 1, 0.2)) +
  scale_fill_viridis_c(option = "mako", begin = 0.3, end = 0.85) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Metric Persistence: H1 vs H2 (2025 NFL Season)",
    x = "Offensive Metric",
    y = "Split-Half Pearson Correlation (r)"
  )
  
```


### Tasks

Repeat the analysis above for the 2022, 2023 and 2024 seasons individually.

```{r}
# Code goes here for 2022
```

```{r}

```


How much do the correlations change from year to year?  Does it make sense to combine the data from multiple years and do those correlations?  Why or why not?




