
---
title: "Introduction to NFL Data Analysis with `nflfastR`"
subtitle: "SPOA 4001/ItCS 5010: American Football Analytics"
author: "Schuckers"
date: "2026-08-16"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-summary: "Show R Code"
    theme: flatly
    highlight-style: github
execute:
  warning: false
  message: false
---

### Overview

In advanced sports analytics, play-by-play data provides the granular foundation needed to evaluate team efficiency, player performance, and tactical decisions. In this lab, we utilize **`nflfastR`**, an R library that provides fast access to cleaned NFL play-by-play data embedded with advanced statistical metrics.


```{r}
library(reticulate)
Sys.setenv(RETICULATE_PYTHON = "C:\\Users\\mschuck1\\AppData\\Local\\Programs\\Python\\Python313\\python.exe")

```


```{python}
print("world")
```


### Environment Setup & Required Libraries

Ensure you have the required packages installed before running this document:

```{r}
# Uncomment if packages are not yet installed:
# install.packages(c("tidyverse", "nflfastR", "nflplotR", "gt", "knitr"))
library(tidyverse)
library(nflfastR)
library(nflplotR)
library(gt)
library(knitr)

# Set global option to avoid scientific notation in printouts
options(scipen = 999)
```

## Starting with Play by Play (pbp) data

In the interest of speed, I've only taken regular season data from the first 4 weeks of the 2025 season.  Actually, I've taken the full 2025 play by play and I've 
```{r}
# Fetch 2025 play-by-play data
raw_pbp_2025 <- nflfastR::load_pbp(2025)

# Filter for Weeks 1-4 regular season scrimmage plays
pbp_w1_4 <- raw_pbp_2025 |>
  filter(
    season_type == "REG",
    week <= 4,
    !is.na(posteam),                  
    # Ensure there is an offensive team
    play_type %in% c("pass", "run") | pass == 1 | rush == 1,
    !is.na(epa)                       # Retain valid EPA values
  )

# Inspect dataset dimensions
cat("Dataset dimensions (rows x cols):", dim(pbp_w1_4), "\n")
```
What variables are available to us
```{r}
names(pbp_w1_4)
```
Only 370 features/variables per play.

The next set of code aggregates all of the play by play data by team.  We get some variables based upon expected points added (EPA) per play.  We do this aggregation by offense and defense then join them back together.

```{r}
#Offensive Aggregation (posteam)
offense_summary <- pbp_w1_4 |>
  group_by(posteam) |>
  summarize(
    off_plays = n(),
    off_epa_per_play = mean(epa, na.rm = TRUE),
    pass_epa = mean(epa[pass == 1], na.rm = TRUE),
    rush_epa = mean(epa[rush == 1], na.rm = TRUE),
    off_success_rate = mean(success, na.rm = TRUE),
    dropback_pct = mean(pass == 1, na.rm = TRUE),
    .groups = "drop"
  )

# Defensive Aggregation (defteam)
defense_summary <- pbp_w1_4 |>
  group_by(defteam) |>
  summarize(
    def_plays = n(),
    def_epa_allowed = mean(epa, na.rm = TRUE),
    def_pass_epa_allowed = mean(epa[pass == 1], na.rm = TRUE),
    def_rush_epa_allowed = mean(epa[rush == 1], na.rm = TRUE),
    def_success_rate_allowed = mean(success, na.rm = TRUE),
    .groups = "drop"
  )

# Join Offense & Defense into a single Master Team Summary
team_summary <- offense_summary |>
  inner_join(defense_summary, by = c("posteam" = "defteam")) |>
  rename(team = posteam) |>
  arrange(desc(off_epa_per_play)) 

# Display top 5 teams by Offensive EPA
head(team_summary, 5) |> kable()

#Display top 6 teams by Defenseive EPA
team_summary<- team_summary |>
  arrange(def_epa_allowed)

head(team_summary, 6) |> kable (digits=3)



```
From the data above, which was the third best team on offensive EPA/play from these data?

Let's make a better table.  

```{r}
#reorder the data for the plot below
team_summary<- team_summary |>
  arrange(desc(off_epa_per_play))

# create a pretty table with some coloring
team_summary |>
  gt() |>
  tab_header(
    title = md("**NFL Team Performance Summary: 2025 (Weeks 1–4)**"),
    subtitle = "Offensive and Defensive Efficiency Metrics Calculated via nflfastR"
  ) |>
  cols_label(
    team = "Team",
    off_plays = "Off. Plays",
    off_epa_per_play = "Off EPA/Play",
    pass_epa = "Pass EPA",
    rush_epa = "Rush EPA",
    off_success_rate = "Off. Success Rate",
    dropback_pct = "Pass Rate",
    def_epa_allowed = "Def EPA/Play",
    def_success_rate_allowed = "Def. Success Rate"
  ) |>
  cols_hide(columns = c(def_plays, def_pass_epa_allowed, def_rush_epa_allowed)) |>
  fmt_number(
    columns = c(off_epa_per_play, pass_epa, rush_epa, def_epa_allowed),
    decimals = 3
  ) |>
  fmt_percent(
    columns = c(off_success_rate, dropback_pct, def_success_rate_allowed),
    decimals = 1
  ) |>
  data_color(
    columns = off_epa_per_play,
    palette = c("#f87171", "#ffffff", "#4ade80")
  ) |>
  tab_source_note(
    source_note = "Data Source: nflfastR | Scrimmage plays only (Weeks 1–4, 2025)"
  )
```

Let's visualize some of these data.

```{r}
# Compute tier averages for reference lines
avg_off_epa <- mean(team_summary$off_epa_per_play, na.rm = TRUE)
avg_def_epa <- mean(team_summary$def_epa_allowed, na.rm = TRUE)

ggplot(team_summary, aes(x = off_epa_per_play, y = def_epa_allowed)) +
  # Quadrant Mean Lines
  geom_hline(yintercept = avg_def_epa, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_vline(xintercept = avg_off_epa, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  # Official NFL Team Logos
  geom_nfl_logos(aes(team_abbr = team), width = 0.055, alpha = 0.9) +
  # Invert Y-axis so lower EPA allowed (better defense) is higher up
  scale_y_reverse() +
  labs(
    title = "2025 NFL Team Tier Breakdown (Weeks 1–4)",
    subtitle = "Offensive Efficiency vs. Defensive Efficiency",
    x = "Offensive EPA per Play (Higher = Better Offense)",
    y = "Defensive EPA per Play Allowed (Lower = Better Defense)",
    caption = "Chart: Senior Data Science | Data: nflfastR"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(color = "gray30", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
```
Note that the plot above has an unusual y-axis.  Why is that?


### Tasks

In sports analytics, plays occurring when the win probability is near 0% or 100% ("garbage time") can skew team metrics.  Modify the filtering logic to exclude plays where win probability < 0.05 or  > 0.95. How does removing garbage time affect the top 3 offensive teams?

#### Task 1
```{r}
# Student Workspace: Filter out garbage time


# TODO: Compute updated team summaries and compare results

```

#### Task 2

Is this the best definition of garbage time?  How might you define garbage time?




## Weekly Data 
Executive Overview & Learning Objectives

Analyzing single-week snapshots introduces high sampling noise. Expanding analysis across multiple full seasons (2022–2025) allows us to measure week-to-week stability, evaluate intra-season variance, and track longitudinal offensive and defensive efficiency.

### Key Objectives
By the end of this lab, students will be able to:
1. **Multi-Year Data Pipeline:** Efficiently import, cache, and clean multi-season play-by-play datasets using `nflfastR`.
2. **Weekly Granular Aggregation:** Construct a panel dataset aggregated at the `(season, week, team)` level.
3. **Time-Series Metric Engineering:** Calculate 3-week rolling averages for Expected Points Added ($\text{EPA}$) to smooth out single-game noise.
4. **Evaluating Metric Autocorrelation:** Quantify the predictive stability of week-to-week efficiency using lag correlations.

---

