
---
title: "Introduction to NFL Data Analysis with `nflfastR`"
subtitle: "SPOA 4001/ItCS 5010: American Football Analytics"
author: "Schuckers"
date: "2026-08-16"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-summary: "Show R Code"
    theme: flatly
    highlight-style: github
execute:
  output: false
  warning: false
  message: false
---

### Overview

In advanced sports analytics, play-by-play data provides the granular foundation needed to evaluate team efficiency, player performance, and tactical decisions. In this lab, we utilize **`nflfastR`**, an R library that provides fast access to cleaned NFL play-by-play data embedded with advanced statistical metrics.

Slight digression to test some python usage in this Quarto markdown file.

```{r}
library(reticulate)
Sys.setenv(RETICULATE_PYTHON = "C:\\Users\\mschuck1\\AppData\\Local\\Programs\\Python\\Python313\\python.exe")

```


```{python}
print("world")
```


### Environment Setup & Required Libraries

Ensure you have the required packages installed before running this document:

```{r}
# Uncomment if packages are not yet installed:
# install.packages(c("tidyverse", "nflfastR", "nflplotR", "gt", "knitr"))
library(tidyverse)
library(nflfastR)
library(nflplotR)
library(gt)
library(knitr)

# Set global option to avoid scientific notation in printouts
options(scipen = 999)
```

## Starting with Play by Play (pbp) data

In the interest of speed, I've only taken regular season data from the first 4 weeks of the 2025 season.  Actually, I've taken the full 2025 play by play and I've 
```{r}
# Fetch 2025 play-by-play data
raw_pbp_2025 <- nflfastR::load_pbp(2025)

# Filter for Weeks 1-4 regular season scrimmage plays
pbp_w1_4 <- raw_pbp_2025 |>
  filter(
    season_type == "REG",
    week <= 4,
    !is.na(posteam),                  
    # Ensure there is an offensive team
    play_type %in% c("pass", "run") | pass == 1 | rush == 1,
    !is.na(epa)                       # Retain valid EPA values
  )

# Inspect dataset dimensions
cat("Dataset dimensions (rows x cols):", dim(pbp_w1_4), "\n")
```
What variables are available to us
```{r}
names(pbp_w1_4)
```
Only 370 features/variables per play.

Let's take a look at the first 6 rows of these data
```{r}
head(pbp_w1_4)
```

The next set of code aggregates all of the play by play data by team.  We get some variables based upon expected points added (EPA) per play.  We do this aggregation by offense and defense then join them back together.

```{r}
#Offensive Aggregation (posteam)
offense_summary <- pbp_w1_4 |>
  group_by(posteam) |>
  summarize(
    off_plays = n(),
    off_epa_per_play = mean(epa, na.rm = TRUE),
    pass_epa = mean(epa[pass == 1], na.rm = TRUE),
    rush_epa = mean(epa[rush == 1], na.rm = TRUE),
    off_success_rate = mean(success, na.rm = TRUE),
    dropback_pct = mean(pass == 1, na.rm = TRUE),
    .groups = "drop"
  )

# Defensive Aggregation (defteam)
defense_summary <- pbp_w1_4 |>
  group_by(defteam) |>
  summarize(
    def_plays = n(),
    def_epa_allowed = mean(epa, na.rm = TRUE),
    def_pass_epa_allowed = mean(epa[pass == 1], na.rm = TRUE),
    def_rush_epa_allowed = mean(epa[rush == 1], na.rm = TRUE),
    def_success_rate_allowed = mean(success, na.rm = TRUE),
    .groups = "drop"
  )

# Join Offense & Defense into a single Master Team Summary
team_summary <- offense_summary |>
  inner_join(defense_summary, by = c("posteam" = "defteam")) |>
  rename(team = posteam) |>
  arrange(desc(off_epa_per_play)) 

# Display top 5 teams by Offensive EPA
head(team_summary, 5) |> kable()

#Display top 6 teams by Defenseive EPA
team_summary<- team_summary |>
  arrange(def_epa_allowed)

head(team_summary, 6) |> kable (digits=3)



```
From the data above, which was the third best team on offensive EPA/play from these data?

Let's make a better table.  

```{r}
#reorder the data for the plot below
team_summary<- team_summary |>
  arrange(desc(off_epa_per_play))

# create a pretty table with some coloring
team_summary |>
  gt() |>
  tab_header(
    title = md("**NFL Team Performance Summary: 2025 (Weeks 1–4)**"),
    subtitle = "Offensive and Defensive Efficiency Metrics Calculated via nflfastR"
  ) |>
  cols_label(
    team = "Team",
    off_plays = "Off. Plays",
    off_epa_per_play = "Off EPA/Play",
    pass_epa = "Pass EPA",
    rush_epa = "Rush EPA",
    off_success_rate = "Off. Success Rate",
    dropback_pct = "Pass Rate",
    def_epa_allowed = "Def EPA/Play",
    def_success_rate_allowed = "Def. Success Rate"
  ) |>
  cols_hide(columns = c(def_plays, def_pass_epa_allowed, def_rush_epa_allowed)) |>
  fmt_number(
    columns = c(off_epa_per_play, pass_epa, rush_epa, def_epa_allowed),
    decimals = 3
  ) |>
  fmt_percent(
    columns = c(off_success_rate, dropback_pct, def_success_rate_allowed),
    decimals = 1
  ) |>
  data_color(
    columns = off_epa_per_play,
    palette = c("#f87171", "#ffffff", "#4ade80")
  ) |>
  tab_source_note(
    source_note = "Data Source: nflfastR | Scrimmage plays only (Weeks 1–4, 2025)"
  )
```

Let's visualize some of these data.

```{r}
# Compute tier averages for reference lines
avg_off_epa <- mean(team_summary$off_epa_per_play, na.rm = TRUE)
avg_def_epa <- mean(team_summary$def_epa_allowed, na.rm = TRUE)

ggplot(team_summary, aes(x = off_epa_per_play, y = def_epa_allowed)) +
  # Quadrant Mean Lines
  geom_hline(yintercept = avg_def_epa, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  geom_vline(xintercept = avg_off_epa, linetype = "dashed", color = "gray50", linewidth = 0.6) +
  # Official NFL Team Logos
  geom_nfl_logos(aes(team_abbr = team), width = 0.055, alpha = 0.9) +
  # Invert Y-axis so lower EPA allowed (better defense) is higher up
  scale_y_reverse() +
  labs(
    title = "2025 NFL Team Tier Breakdown (Weeks 1–4)",
    subtitle = "Offensive Efficiency vs. Defensive Efficiency",
    x = "Offensive EPA per Play (Higher = Better Offense)",
    y = "Defensive EPA per Play Allowed (Lower = Better Defense)",
    caption = "Chart: Senior Data Science | Data: nflfastR"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(color = "gray30", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
```
Note that the plot above has an unusual y-axis.  Why is that?


## Tasks

In sports analytics, plays occurring when the win probability is near 0% or 100% ("garbage time") can skew team metrics.  Modify the filtering logic to exclude plays where win probability < 0.05 or  > 0.95. How does removing garbage time affect the top 3 offensive teams?

### Task 1
```{r}
# Student Workspace: Filter out garbage time


# TODO: Compute updated team summaries and compare results

```

### Task 2

Is this the best definition of garbage time?  How might you define garbage time?




## Drive summary data 


So we are going back to the play by play data from 2025 and now we are going to 
look at some drive summaries.  
```{r}
pbp_clean <- pbp_w1_4 %>%
  filter(
    !is.na(drive),
    !is.na(epa),
    play == 1,                       # Excludes non-play events / dead-ball fouls
    play_type %in% c("run", "pass")  # Restricts to offensive play attempts
  )
```

Next we group the data and since drives are numbered from the beginning of each game, we have to group by game and drive.


```{r}
drive_level_data <- pbp_clean %>%
  group_by(game_id, drive, posteam, defteam) %>%
  summarize(
    drive_total_epa = sum(epa, na.rm = TRUE),
    drive_mean_epa  = mean(epa, na.rm = TRUE),
    play_count      = n(),
    drive_result    = first(fixed_drive_result),
    .groups = "drop"
  )
```

Next take the drive level data and get some offensive and defensive summaries
```{r}
offense_drive_summary <- drive_level_data %>%
  filter(!is.na(posteam)) %>%
  group_by(posteam) %>%
  summarize(
    drives           = n(),
    mean_drive_epa   = mean(drive_total_epa, na.rm = TRUE),
    median_drive_epa = median(drive_total_epa, na.rm = TRUE),
    total_off_epa    = sum(drive_total_epa, na.rm = TRUE),
    avg_plays_per_dr = mean(play_count, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_drive_epa))

head(offense_drive_summary, 6) |> kable (digits=3)

defense_drive_summary <- drive_level_data %>%
  filter(!is.na(defteam)) %>%
  group_by(defteam) %>%
  summarize(
    drives_def        = n(),
    mean_epa_allowed  = mean(drive_total_epa, na.rm = TRUE),
    median_epa_allow  = median(drive_total_epa, na.rm = TRUE),
    total_def_epa     = sum(drive_total_epa, na.rm = TRUE),
    avg_plays_allowed = mean(play_count, na.rm = TRUE)
  ) %>%
  arrange(mean_epa_allowed) # Lower/negative EPA allowed indicates higher defensive

head(defense_drive_summary, 6) |> kable (digits=3)
```
### Task 3

Find another metric, not EPA, and calculate the summaries (mean, median, sum) for each drive in the *raw_pbp_2025* dataset.  Then do it for only pass plays.  Which team was the best on your metric for all plays? for only pass plays?



## College Football Data

Below we do some practice with loading, summarizing and visualizing the college football data.

Using `cfbfastr::load_cfb_pbp()` fetches pre-processed play-by-play data directly without requiring API key authentication. We aggregate play-level yardage to quantify offensive and defensive Yards Per Play (YPP) across the 2025 season.

```{r}
#| label: setup-and-load

library(cfbfastR)
library(ggrepel)

# Load full 2024 play-by-play dataset
pbp_2025 <- load_cfb_pbp(2025) |>
   filter(home_team_division=="fbs" & away_team_division=="fbs")|>
  filter(rush == 1 | pass == 1, !is.na(yards_gained)) 
```

Note pbp for all college is a big file and includes teams from FBS, FCS, D2 and D3.  We're only interested in FBS v FBS games today.

Below we get the yards per play gained for offenses.  

```{r}
# Compute offensive YPP by team
off_ypp <- pbp_2025 |>
  group_by(team = offense_play, conference = offense_conference) |>
  summarize(
    off_plays = n(),
    off_ypp = mean(yards_gained),
    .groups = "drop"
  ) |>
  arrange(desc(off_ypp)) 

# Lets look at who is `best` on this metric
head(off_ypp,15)|>kable(digits=2)
```

Below we get the yards per play gained against defenses.  

```{r}

# Compute defensive YPP allowed by team
def_ypp <- pbp_2025 |>
  group_by(team = defense_play) |>
  summarize(
    def_plays = n(),
    def_ypp = mean(yards_gained),
    .groups = "drop"
  )|>
  arrange(def_ypp) 

# Lets look at who is best
head(def_ypp,15)|>kable(digits=2)

```

Below we join and classify the teams into tiers.  

```{r}

# Join metrics and classify into performance tiers
all_ypp <- off_ypp |>
  inner_join(def_ypp, by = "team") |>
  mutate(
    net_ypp = off_ypp - def_ypp,
    tier = case_when(
      off_ypp > mean(off_ypp) & def_ypp < mean(def_ypp) ~ "Elite",
      off_ypp > mean(off_ypp) & def_ypp >= mean(def_ypp) ~ "High Offense / Low Defense",
      off_ypp <= mean(off_ypp) & def_ypp < mean(def_ypp) ~ "Low Offense / High Defense",
      TRUE ~ "Rebuilding"
    )
  ) |>
  arrange(desc(net_ypp))
```

Print out the data in a nice format.

```{r}

# Render structured output table
all_ypp |>
  select(team, off_ypp, def_ypp, net_ypp, tier) |>
  gt() |>
  tab_header(
    title = "2025 FBS Yards Per Play Summary",
    subtitle = "Evaluated on all rushing and passing attempts"
  ) |>
  cols_label(
    team = "Team",
    off_ypp = "Offensive YPP",
    def_ypp = "Defensive YPP",
    net_ypp = "Net YPP Margin",
    tier = "Performance Tier"
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 2)
```

Now a slick visualization with 'repel' labels to make things readable.

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(all_ypp, aes(x = off_ypp, y = def_ypp)) +
  geom_vline(xintercept = mean(all_ypp$off_ypp), linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = mean(all_ypp$def_ypp), linetype = "dashed", color = "gray50") +
  geom_point(aes(color = tier), size = 3) +
  geom_text_repel(aes(label = team), size = 3.5, max.overlaps = 15) +
  scale_y_reverse() + 
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "2025 Team Yards Per Play Matrix",
    subtitle = "Top-Right quadrant indicates above-average offensive and defensive performance",
    x = "Offensive YPP (Higher is Better)",
    y = "Defensive YPP Allowed (Lower is Better)",
    color = "Tier",
    caption = "Data: cfbfastr"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

### Task 4

Repeat the analysis above for another metric.  This time only include data from games where the score differential between the two teams was less than 21.

```{r}


```


## Using your API key

In the code below put your key within the quotes
```{r echo=FALSE}
Sys.setenv(CFBD_API_KEY = "")
```

```{r}
# Verify API key availability
if (Sys.getenv("CFBD_API_KEY") == "") {
  stop("CFBD_API_KEY environment variable not detected.")
}

# Load live API endpoints: 2025 Team Talent Ratings & Game Results
talent_raw <- cfbd_team_talent(year = 2025) |> rename("team"="school")
games_raw  <- cfbd_game_info(year = 2025, season_type = "regular")
```

Get wins for each team in 2025

```{r}


# Calculate regular season wins per team
home_wins <- games_raw |> 
  filter(!is.na(home_points), home_points > away_points) |> 
  count(team = home_team, name = "wins")

away_wins <- games_raw |> 
  filter(!is.na(away_points), away_points > home_points) |> 
  count(team = away_team, name = "wins")

total_wins <- bind_rows(home_wins, away_wins) |> 
  group_by(team) |> 
  summarize(wins = sum(wins), .groups = "drop")
```



```{r}
# Clean talent scores, join wins, and extract regression residuals
df_analytics <- talent_raw |> 
  transmute(
    team,
    talent_score = as.numeric(talent)
  ) |> 
  inner_join(total_wins, by = "team") |> 
  drop_na()
```


```{r}
# Fit OLS model to derive expected wins and residual performance
model <- lm(wins ~ talent_score, data = df_analytics)

df_analytics <- df_analytics |> 
  mutate(
    expected_wins = predict(model),
    win_residual  = wins - expected_wins,
    efficiency   = if_else(win_residual >= 0, "Overperforming", "Underperforming")
  ) |> 
  arrange(desc(win_residual))

# Tabulate top 5 overperforming teams relative to talent
df_analytics |> 
  slice_head(n = 5) |> 
  gt() |> 
  tab_header(
    title = "Top 5 Talent Overperformers (2024)",
    subtitle = "Teams exceeding expected win totals based on 247Sports Composite Talent"
  ) |> 
  cols_label(
    team = "Team",
    talent_score = "Talent Score",
    wins = "Actual Wins",
    expected_wins = "Expected Wins",
    win_residual = "Win Surplus",
    efficiency = "Status"
  ) |> 
  fmt_number(columns = c(talent_score, expected_wins, win_residual), decimals = 2)

```


Make a visualization of talent vs Wins with the regression line on the plot

```{r}
#| label: visualization
#| fig-width: 9
#| fig-height: 6

ggplot(df_analytics, aes(x = talent_score, y = wins)) +
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "gray80", linetype = "dashed") +
  geom_point(aes(color = win_residual), size = 3.5, alpha = 0.85) +
  geom_text_repel(
    aes(label = team), 
    size = 3, 
    max.overlaps = 12,
    box.padding = 0.3
  ) +
  scale_color_gradient2(
    low = "#d73027", 
    mid = "#ffffbf", 
    high = "#1a9850", 
    midpoint = 0,
    name = "Win Surplus"
  ) +
  labs(
    title = "2025 FBS Talent Composite vs. Regular Season Wins",
    subtitle = "Dashed line indicates baseline OLS expected wins given roster talent",
    x = "247Sports Team Talent Score",
    y = "Actual Regular Season Wins",
    caption = "Data fetched via live API: cfbfastr"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

