---
title: "College Football Team Ratings (2022–2025)"
author: "Schuckers"
format: 
  html:
    toc: true
    theme: cosmo
execute:
  output: false
  warning: false
  message: false
---


### Overview

This analysis fetches College Football (CFB) game scores from 2022 through 2025 using `cfbfastR` and calculates three rating models from first principles without external rating packages:



```{r setup}
# Install required packages if missing:
# install.packages(c("cfbfastR", "dplyr", "tidyr", "purrr", "knitr"))

library(cfbfastR)
library(dplyr)
library(tidyr)
library(purrr)
library(knitr)
```

Data Import & Preprocessing
```{r}

seasons <- 2022:2025

raw_games <- load_cfb_schedules(seasons)

# Clean and filter for completed games with valid scores
games_clean <- raw_games %>%
  filter(
    completed == TRUE,
    !is.na(home_points),
    !is.na(away_points),
    away_division=="fbs",
    season_type %in% c("regular")
  ) %>%
  select(season, week, home_team, away_team, home_points, away_points) %>%
  arrange(season, week)
```

Here I've given you functions *from scratch* meaning that the code below
does the calculations.  Note that there are packages/libraries in R and Python for doing all of these but you don't get to see the raw calculations.
```{r}

# 1. Scratch Elo Calculation
calc_elo_scratch <- function(df, k_val = 20, init_rating = 1500) {
  teams <- unique(c(df$home_team, df$away_team))
  ratings <- setNames(rep(init_rating, length(teams)), teams)
  
  for (i in seq_len(nrow(df))) {
    home <- df$home_team[i]
    away <- df$away_team[i]
    
    r_home <- ratings[home]
    r_away <- ratings[away]
    
    # Expected win probability for home team
    e_home <- 1 / (1 + 10^((r_away - r_home) / 400))
    
    # Actual score outcome (1 = Win, 0.5 = Tie, 0 = Loss)
    if (df$home_points[i] > df$away_points[i]) {
      s_home <- 1
    } else if (df$home_points[i] < df$away_points[i]) {
      s_home <- 0
    } else {
      s_home <- 0.5
    }
    
    # Update ratings
    ratings[home] <- r_home + k_val * (s_home - e_home)
    ratings[away] <- r_away + k_val * ((1 - s_home) - (1 - e_home))
  }
  
  data.frame(
    team = names(ratings),
    Elo = round(unname(ratings), 1),
    stringsAsFactors = FALSE
  )
}
```

Here's the Bradley-Terry formulation via the Logistic Regression.  Note that we start with a matrix of zeroes with a column for each team and a row for each game.  We then fill in with $1$'s and $-1$'s for the home team and the away team respectively.  

```{r}
# Scratch Bradley-Terry Calculation (Logistic Regression)
calc_bt_scratch <- function(df) {
  # Remove ties if any exist
  df_bt <- df %>% filter(home_points != away_points)
  
  teams <- unique(c(df_bt$home_team, df_bt$away_team))
  n_games <- nrow(df_bt)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df_bt$home_team[i]] <- 1
    X[i, df_bt$away_team[i]] <- -1
  }
  
  y <- ifelse(df_bt$home_points > df_bt$away_points, 1, 0)
  
  # Fit logistic regression model without intercept
  fit <- glm(y ~ 0 + X, family = binomial(link = "logit"))
  
  bt_vec <- coef(fit)
  names(bt_vec) <- gsub("^X", "", names(bt_vec))
  bt_vec[is.na(bt_vec)] <- 0  # Replace unestimated reference level with 0
  
  # Center ability parameters around zero
  bt_centered <- bt_vec - mean(bt_vec, na.rm = TRUE)
  
  data.frame(
    team = names(bt_centered),
    BradleyTerry = round(unname(bt_centered), 3),
    stringsAsFactors = FALSE
  )
}
```

We do the same for the Simple Rating System (SRS) Calculation (Linear Model)
```{r}
calc_srs_scratch <- function(df) {
  teams <- unique(c(df$home_team, df$away_team))
  n_games <- nrow(df)
  
  # Construct design matrix: Home (+1), Away (-1)
  X <- matrix(0, nrow = n_games, ncol = length(teams))
  colnames(X) <- teams
  
  for (i in seq_len(n_games)) {
    X[i, df$home_team[i]] <- 1
    X[i, df$away_team[i]] <- -1
  }
  
  margin <- df$home_points - df$away_points
  fit <- lm(margin ~ 0 + X)
  
  srs_vec <- coef(fit)
  names(srs_vec) <- gsub("^X", "", names(srs_vec))
  srs_vec[is.na(srs_vec)] <- 0
  
  srs_centered <- srs_vec - mean(srs_vec, na.rm = TRUE)
  
  data.frame(
    team = names(srs_centered),
    SRS = round(unname(srs_centered), 2),
    stringsAsFactors = FALSE
  )
}
```


Below is a function for calculating the ratings by season for each of the methods above.  
```{r}

compute_season_ratings <- function(season_df) {
  year <- season_df$season[1]
  
  elo_df <- calc_elo_scratch(season_df)
  bt_df  <- calc_bt_scratch(season_df)
  srs_df <- calc_srs_scratch(season_df)
  
  elo_df %>%
    full_join(bt_df, by = "team") %>%
    full_join(srs_df, by = "team") %>%
    mutate(season = year) %>%
    arrange(desc(SRS))
}

# Execute loop across split season datasets
ratings_by_season <- games_clean %>%
  group_split(season) %>%
  map(compute_season_ratings) %>%
  bind_rows()
```

Top 10 Rankings by Season

```{r}

for (yr in sort(unique(ratings_by_season$season))) {
  cat("\n### Season:", yr, "\n\n")
  
  top_rankings <- ratings_by_season %>%
    filter(season == yr) %>%
    arrange(desc(SRS)) %>%
    slice_head(n = 10) %>%
    mutate(Rank = row_number()) %>%
    select(
      Rank,
      Team = team,
      `SRS Rating` = SRS,
      `Elo Rating` = Elo,
      `Bradley-Terry` = BradleyTerry
    )
  
  print(
    kable(
      top_rankings,
      format = "markdown",
      align = c("c", "l", "r", "r", "r")
    )
  )
  cat("\n")
}

```


